---
title: "STATS306_WN23_Lab15"
author: "Zehua Wang"
date: "2023-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# STATS306 Lab15 Shiny

## Introduction

Shiny is an R package that makes it easy to build interactive web applications (apps)
straight from R. Specifically, users can build web applications without needing 
to know HTML, CSS, or JavaScript. Shiny R provides a framework for building web-based
dashboards and applications that allow users to interact with data and perform 
analyses in real-time, making it an ideal tool for data exploration and visualization.

```{r}
# install.packages("shiny")
library(shiny)
```

Let's take a look at some examples

```{r}
runExample("01_hello")
```

In this example, we draw histogram of the wait time between eruptions for the Old 
Faithful geyser in Yellowstone National Park in dataset faith. The interactive bar 
allows us to change the number of bins of histogram.

```{r}
runExample("03_reactivity")
```

In this example, we change the caption of summary table, choose different dataset, 
and also the number of observation for preview.

```{r}
runExample("04_mpg")
```

In this example, we generate boxplots between mpg and categorical variables including
cylinders, transmission, and gear. We also have an option for the outlier.

In this lab, we will learn how to filter and visualize data using `ncaa` dataset. 
Before we moving on, lets clean the dataset.

```{r}
library(tidyverse)
library(lubridate)
ncaa <- read_csv("https://datasets.stats306.org/ncaa.csv.gz", show_col_types = FALSE)
```

## Structure

The structure of a Shiny App includes:

-   a user interface object
-   a server function
-   a call to `shinyApp` function

The user interface (`ui`) object controls the layout and appearance of your app. 
The `server` function contains the instructions that your computer needs to build 
your app. Finally the `shinyApp` function creates Shiny app objects from an 
explicit UI/server pair.

Sample structure code:

```{r}
# Define UI ----
ui <- fluidPage(
  
)

# Define server logic ----
server <- function(input, output) {
  
}

# Run the app ----
shinyApp(ui = ui, server = server)

```

Here is a useful shiny [cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/shiny.pdf)

### User interface object

In Shiny R, user interface (UI) is the front-end part of a Shiny app, which defines 
the layout and appearance of the app in the web browser. The UI part is created 
using a set of functions in `shiny` library. It can be customized to include a wide 
range of input widgets, output elements, and HTML components.

Shiny uses `fliudPage` to creates a UI that includes title panel and sidebar layout, 
which includes sidebar panel and main panel. Run the following code and see the layout.

```{r}
# Define UI ----
ui <- fluidPage(
  titlePanel("title panel"),

  sidebarLayout(
    sidebarPanel("sidebar panel"),
    mainPanel("main panel")
  )
)

# Define server logic ----
server <- function(input, output) {
  
}

# Run the app ----
shinyApp(ui = ui, server = server)
```

Shiny uses widgets as UI element that allows users to interact with the application. 
It could be an input widget like a text box, a dropdown menu, a slider, or a button; 
or an output widget like a plot, a table, or a text display.

Here are some commonly used input widgets:

-   `textInput()`: creates a single-line text input field.
-   `textareaInput()`: creates a multi-line text input field.
-   `numericInput()`: creates a numeric input field.
-   `selectInput()`: creates a dropdown menu to select a single item from a list of choices.
-   `selectizeInput()`: creates a searchable dropdown menu to select a single item from a list of choices.
-   `checkboxInput()`: creates a single checkbox.
-   `checkboxGroupInput()`: creates a group of checkboxes.
-   `radioButtons()`: creates a group of radio buttons to select a single item from a list of choices.
-   `sliderInput()`: creates a slider to select a numeric value within a range.

You can check the usage of shiny widgets in the [gallery](https://shiny.rstudio.com/gallery/widget-gallery.html).

Let's creates some widgets for filtering NCAA data.

Use `sliderInput()` to select the range of year. We need to specify the `inputId`, `label` of widget, `min` and `max`. `value` used as default input value.

```{r}
sliderInput(inputId="years", label = "Year Range", min = 1985,
                           max = 2016, value = c(1990, 2010))
```

Use `selectInput()` to select teams. Similar to `sliderInput()`, we need to specify `choices`. `selected` used as default input value.

```{r}
selectInput("team2", label = "Select Team", choices = unique(c(ncaa$Winner,  ncaa$Loser)),
                           selected = "Michigan")
```

However, it takes time to find the team in drop down menu when there are many teams. 
In this case, you may consider to use `selectizeInput()`. It allows user to select one 
or multiple options from a list of choices. It creates an auto-complete search box 
that filters the choices based on the user input.

```{r}
selectizeInput(
   inputId = "team1", 
   label = "Search Team",
   choices = unique(c(ncaa$Winner,  ncaa$Loser)),
   options = list(
     placeholder = "Search Team",
     maxItems = '1',
     onDropdownOpen = I("function($dropdown) {if (!this.lastQuery.length) {this.close(); this.settings.openOnFocus = false;}}"),
     onType = I("function (str) {if (str === \"\") {this.close();}}")
   ),
   selected = "Michigan"
 )
```

We create two ways to choose a team. Use `radioButtons()` to select the team name 
from either `search team` or `select team`. The default choice is `search team`.

```{r}
radioButtons("team_choice", label = "Which team",
                            choices = list("Search Team" = "T1", "Select Team" = "T2"), 
                            selected = "T1")

```

Use `numericInput()` to select the number of observations or rows in the summarize table.
The default value is 5.

```{r}
numericInput("num_obs", label = "Number of Observations", value = 5)
```

Use `checkboxGroupInput()` to select the regions and round. We choose to select all by default.

```{r}
checkboxGroupInput("region", label = "Region", choices = unique(ncaa$Region),
                                  selected = unique(ncaa$Region))
checkboxGroupInput("round", label = "Round", choices = unique(ncaa$Round),
                                  selected = unique(ncaa$Round))
```

Using `fluidRow` and `column` to arrange the widgets in `sidebarPanel`. Then, we create a widget input UI.

```{r}
# Define UI ----
ui <- fluidPage(
  titlePanel("NCAA Data"),
  
  sidebarLayout(
    
    sidebarPanel(
      helpText("Summarize and visualize the NCAA data from 1985 to 2016"), # add brief explanation for app usage 
      
      fluidRow(
        column(12,
               # Year range
               sliderInput("years", label = "Year Range", min = 1985,
                           max = 2016, value = c(1990, 2010))
        )
      ),
      
      fluidRow(
        # Team select in two ways
        column(6, # column width must be between 1 and 12
               selectizeInput(
                 inputId = "team1", 
                 label = "Search Team",
                 choices = unique(c(ncaa$Winner,  ncaa$Loser)),
                 options = list(
                   placeholder = "Search Team",
                   maxItems = '1',
                   onDropdownOpen = I("function($dropdown) {if (!this.lastQuery.length) {this.close();
                                      this.settings.openOnFocus = false;}}"),
                   onType = I("function (str) {if (str === \"\") {this.close();}}")
                 ),
                 selected = "Michigan"
               )
        ),
        
        column(6, 
               selectInput("team2", label = "Select Team", choices = unique(c(ncaa$Winner,  ncaa$Loser)),
                           selected = "Michigan")
        ),
        
        column(4,
               radioButtons("team_choice", label = "Which team",
                            choices = list("Search Team" = "T1", "Select Team" = "T2"), 
                            selected = "T1")
        ),
        
        column(7,
               numericInput("num_obs", label = "Number of Observations", value = 5))
      ),
      
      fluidRow(
        # Select Region
        column(6,
               checkboxGroupInput("region", label = "Region", choices = unique(ncaa$Region),
                                  selected = unique(ncaa$Region))
        ),
        # select Round
        column(6,
               checkboxGroupInput("round", label = "Round", choices = unique(ncaa$Round),
                                  selected = unique(ncaa$Round))
        )
      )
      
    ),
    
    mainPanel("mainPanel")
    
  )
)

# Define server logic ----
server <- function(input, output) {
  
}

# Run the app ----
shinyApp(ui = ui, server = server)
```

### Server function

In Shiny R, the server function is where you define the logic of your application. 
It contains reactive expressions and functions that control how the user input is 
processed and how the output is rendered. The server function is executed every time 
the user interacts with the application.

Here, we use `reactive` function to create reactive expression. Reactive expression
defines between inputs and outputs in Shiny app. When one of the inputs changes, 
any reactive expressions that depend on that input will be invalidated and re-evaluated,
and any outputs that depend on those reactive expressions will be updated accordingly.

There are some commonly used output widgets for reactive expressions:

- `renderPlot()`: creates a reactive plot.
- `renderTable()`: creates a reactive table.
- `renderText()`: creates a reactive text display.

We will generate a table for filtered dataset using `renderTable()`, and a 
plot of team advance in March Madness similar to homework 8 Problem 1 using
`renderPlot()`.


We generate a filtered table based on `input` from widgets under reactive environment.
Pass the reactive expression to `renderTable()` and assign it to output with `outputId`
`summarize_data`. We generally display the output in `mainPanel`

Code for filtered table:
```{r}
server <- function(input, output) {
  
  filter_tb <-  reactive({
    team <- if (input$team_choice == "T1") input$team1 else input$team2
    tb <- ncaa %>% mutate(Year = year(mdy(Date))) %>%
      filter(Winner == team | Loser == team) %>%
      filter(Year >= input$years[1] & Year <= input$years[2]) %>%
      filter(Region %in% input$region) %>%
      filter(Round %in% input$round)
    return(tb[1:min(dim(tb)[1], input$num_obs),])
  })
  output$summarize_data <- renderTable(filter_tb())
}
```

Similarly, we generate a plot based on `input` from widgets under reactive environment.
Pass the reactive expression to `renderPlot()` and assign it to output with `outputId`
`plot`.

Code for plot:
```{r}
server <- function(input, output) {
  plot <- reactive({
    team <- if (input$team_choice == "T1") input$team1 else input$team2
    RoundLevel = c(unique(ncaa$Round)[7], unique(ncaa$Round)[1:6])
    data <- ncaa %>% mutate(Year = year(mdy(Date))) %>%
      mutate(Round = factor(Round, levels = RoundLevel, ordered=T)) %>%
      filter(Winner == team | Loser == team) %>%
      filter(Year >= input$years[1] & Year <= input$years[2]) %>%
      mutate(champion = Round == "National Championship" & Winner == team) %>%
      group_by(Year) %>% summarize(max.round = max(Round),
            champ = ifelse(as.logical(max(champion)), "Won Championship", NA))
    plot <- data %>% ggplot(aes(x=Year, y=max.round, fill=champ)) + geom_col(orientation="x") +
      scale_y_discrete(drop=F) +
      labs(x="Year", y="Furthest round reached", title=paste("March Madness results: ", team),
           fill="") + theme_classic() + scale_fill_manual(values = "#ee3333", limits = "Won Championship")
    return(plot)
  })
  output$plot <- renderPlot(plot())
}
```


Commonly used display widgets are:

- `plotOutput()`: creates a space to display a plot.
- `tableOutput()`: creates a space to display a table.
- `textOutput()`: creates a space to display text.

Remember, we need to display the outputs. Add the codes in `mainPanel` in UI.

```{r}
mainPanel(
  tableOutput("summarize_data"),
  plotOutput("plot")
)
```

### Call shinyApp

The `shinyApp` function creates Shiny app objects from an explicit UI/server pair. 
It is called at the end of the app to run the Shiny application,

```{r}
shinyApp(ui = ui, server = server)
```

Here is the complete code for NCAA Data shiny app:
```{r}
library(shiny)
library(tidyverse)
library(lubridate)

ncaa <- read_csv("https://datasets.stats306.org/ncaa.csv.gz")

ui <- fluidPage(
  titlePanel("NCAA Data"),
  
  sidebarLayout(
    
    sidebarPanel(
      # add brief explanation for app usage 
      helpText("Summarize and visualize the NCAA data from 1985 to 2016"), 
      
      fluidRow(
        column(12,
               # Year range
               sliderInput("years", label = "Year Range", min = 1985,
                           max = 2016, value = c(1990, 2010))
        )
      ),
      
      fluidRow(
        # Team select in two ways
        column(6, # column width must be between 1 and 12
               selectizeInput(
                 inputId = "team1", 
                 label = "Search Team",
                 choices = unique(c(ncaa$Winner,  ncaa$Loser)),
                 options = list(
                   placeholder = "Search Team",
                   maxItems = '1',
                   onDropdownOpen = I("function($dropdown) {if (!this.lastQuery.length) {this.close();
                                      this.settings.openOnFocus = false;}}"),
                   onType = I("function (str) {if (str === \"\") {this.close();}}")
                 ),
                 selected = "Michigan"
               )
        ),
        
        column(6, 
               selectInput("team2", label = "Select Team", choices = unique(c(ncaa$Winner,  ncaa$Loser)),
                           selected = "Michigan")
        ),
        
        column(4,
               radioButtons("team_choice", label = "Which team",
                            choices = list("Search Team" = "T1", "Select Team" = "T2"), 
                            selected = "T1")
        ),
        
        column(7,
               numericInput("num_obs", label = "Number of Observations", value = 5))
      ),
      
      fluidRow(
        # Select Region
        column(6,
               checkboxGroupInput("region", label = "Region", choices = unique(ncaa$Region),
                                  selected = unique(ncaa$Region))
        ),
        # select Round
        column(6,
               checkboxGroupInput("round", label = "Round", choices = unique(ncaa$Round),
                                  selected = unique(ncaa$Round))
        )
      )
      
    ),
    
    mainPanel(
      tableOutput("summarize_data"),
      plotOutput("plot")
    )
    
  )
)

# Define server logic ----
server <- function(input, output) {
  
  filter_tb <-  reactive({
    team <- if (input$team_choice == "T1") input$team1 else input$team2
    tb <- ncaa %>% mutate(Year = year(mdy(Date))) %>%
      filter(Winner == team | Loser == team) %>%
      filter(Year >= input$years[1] & Year <= input$years[2]) %>%
      filter(Region %in% input$region) %>%
      filter(Round %in% input$round)
    return(tb[1:min(dim(tb)[1], input$num_obs),])
  })
  
  plot <- reactive({
    team <- if (input$team_choice == "T1") input$team1 else input$team2
    RoundLevel = c(unique(ncaa$Round)[7], unique(ncaa$Round)[1:6])
    data <- ncaa %>% mutate(Year = year(mdy(Date))) %>%
      mutate(Round = factor(Round, levels = RoundLevel, ordered=T)) %>%
      filter(Winner == team | Loser == team) %>%
      filter(Year >= input$years[1] & Year <= input$years[2]) %>%
      mutate(champion = Round == "National Championship" & Winner == team) %>%
      group_by(Year) %>% summarize(max.round = max(Round),
            champ = ifelse(as.logical(max(champion)), "Won Championship", NA))
    plot <- data %>% ggplot(aes(x=Year, y=max.round, fill=champ)) + geom_col(orientation="x") +
      scale_y_discrete(drop=F) +
      labs(x="Year", y="Furthest round reached", title=paste("March Madness results: ", team),
           fill="") + theme_classic() + scale_fill_manual(values = "#ee3333", limits = "Won Championship")
    return(plot)
  })
  
  output$summarize_data <- renderTable(filter_tb())
  output$plot <- renderPlot(plot())
}

# Run the app ----
shinyApp(ui = ui, server = server)
```